---
title: "M2RParallelSort"
author: "Olaniyan Folajmi and Gabriele Degola"
date: "11/22/2021"
output:
  html_document: default
---

```{r}
library(ggplot2) # Data visualization
library(readr) # CSV file I/O, e.g. the read_csv function
library(readr)
library(Rtsne)
library(doParallel)
library(benchmarkme)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Parallel Quicksort Analysis
In this work, we will like to compare the performance of the sequential quick sort algorithm to its sequential counterpart. 
While the parallel version of the quicksort algorithm is slower for smaller input sizes, for sufficiently large input sizes,
parallel quicksort outperforms the sequential version. 

## Machine Specification
### Machine CPU
```{r}
cat("Machine:     "); print(get_cpu()$model_name)
```

### Number of cores
```{r}
cat("Num cores:   "); print(detectCores(all.tests = TRUE, logical = FALSE)/2)
```

### Number of threads
```{r}
cat("Num threads: "); print(detectCores(logical = TRUE))
```

### Machine RAM
```{r}
cat("RAM:         "); print (get_ram()); cat("\n")
```
